Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-25 ms.tex
mwp edits

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Daniel S. Falster$^1$, Richard G. FitzJohn$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
\affiliation{
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
University of New South Wales, Sydney, NSW 2052, Australia\\
$^2$ Imperial College, London, United Kingdom\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
}
\date{}

@@

The sharing and re-use of data has become a cornerstone of modern science.  To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.  Thus far, however, this framework does not account for on-going scientific progress: for many important problems, epitomised by problems that lead to one or more meta-analyses, datasets will continue to grow with time---more data points will be added, and new and more useful data structures will be created.  Essentially this means that datasets, like science in general, progress with time.  We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.  In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets. We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.

@@

As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.  These datasets are constantly evolving, i.e. never "finished", and may be the subject of repeated analysis or meta-analysis.

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.

Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

@@

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it \emph{might} change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

@@

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.  Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call. The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.

@@

%work on this paragraph more
The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.  Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but, while there are cases where data curation may be done by a large institution, the current structure of most scientific research is very distributed, with a few research groups across the world working on similar questions.  Moreover, curation of data on particular questions requires, in many cases, a great deal of expertise in the field, so the experts in the field need to be involved in data curation. However, individual research groups have different capabilities compared to institutional data curators: at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such, we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.

@@

One of the greatest benefits of using a cloud--based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy.

@@

Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.

>>>>

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Daniel S. Falster$^1$, Richard G. FitzJohn$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{1,*}$}
\affiliation{
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
University of New South Wales, Sydney, NSW 2052, Australia\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
}
\date{}

@@

The sharing and re-use of data has become a cornerstone of modern science.  To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.  Thus far, however, this framework does not accommodate on-going scientific progress: for many important problems, particularly those suitable for meta-analyses, datasets will continue to grow with time --- more data points will be added, and new and more useful data structures will be created.  Essentially this means that datasets, like science in general, improve with time.  We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.  In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets. We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.

@@

As is evidenced in the advent of this journal \citep{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \citep[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.  These datasets are constantly evolving, i.e. they are never "finished", and may be the subject of repeated analysis or meta-analysis.

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while --- to ensure reproducibility --- past versions of the dataset would remain accessible.

Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

@@

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

Currently there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

@@

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so it is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.  Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call. The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solution. But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions and involve standard data types (e.g., sequences). But for an individual research group with expertise on a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.

@@

One of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy.

@@

Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data. Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
